Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus (Extended Version)

### 2025-04-08 paper/paper.tex
arxiv xcolor weirdness
spelling + arxiv bibtex

% -*- fill-column: 80; -*-
\newif\ifarxiv
\arxivtrue

@@

%% Bibliography style
\bibliographystyle{ACM-Reference-Format}
%% Citation style
%% Note: author/year citations are required for papers published as an
%% issue of PACMPL.
\citestyle{acmauthoryear}   %% For author/year citations
%% \citestyle{acmnumeric}
\usepackage{mathpartir}
\usepackage{tikz-cd}
\usepackage{lipsum}
\usepackage{enumitem}
\usepackage{wrapfig}
\usepackage{fancyvrb}
\usepackage{xspace}
\usepackage[dvipsnames]{xcolor}
\usepackage{thmtools} % needed for sane cleveref naming of theorems
\usepackage{hyperref}
\usepackage[capitalise]{cleveref}

@@

\title{Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus}

@@

\ifarxiv\else {\textbf{Extended version:} An extended version of this paper is available in \citet{extendedLambekD}.}\fi

@@

In addition to these type-theoretic principles, we need two additional
axioms that do not generally hold in systems based on linear
logic. First, we need that additive conjunction \emph{distributes}
over additive disjunction --- e.g., in the finitary case that $0 \& A
\cong 0$ and $(A + B) \& C \cong (A \& C) + (B \& C)$. More generally, we assume
\cref{ax:dist}.
% Comment from shepherd
% - The pagebreaks and figures make the explanation of axiom 3.1 a
% little confusing to read (with the text "In its most general form,"
% separated from the axiom it is intended to introduce by a pagebreak
% and a figure). You could fiddle with the layout a bit, or just revise
% that sentence to "More generally, we assume Axiom 3.1.", or similar.
%
% - If I understand it correctly, axiom 3.1 itself is using "e" as a variable name, which is confusing since it's also used as a metavariable for terms - I suggest picking a different variable name.
\begin{axiom}[Distributivity]
\label{ax:dist}
For any $A : (x : X) \to Y(x) \to L$, the definable distribuitivity function
%%%%%%%%%% An attempt to inline
% \(
%  {\LinSigTy{f}{\left(\PiTy{x}{X}{Y(x)}\right)}{\LinPiTy{x}{X}{A\,x\,(f\,x)}} \lto
% \LinPiTy{x}{X}{\LinSigTy{y}{Y(x)}{A\,x\,y}} \right)}
% \)
\(
 \lamblto{b}{\letin{\sigma\,f\,b'}{b}{\withlamb{x}{b'\,.\pi\,x}}} : \LinSigTyLimit{f}{\PiTy{x}{X}{Y(x)}}{\LinPiTyLimit{x}{X}{A\,x\,(f\,x)}} \lto
\LinPiTyLimit{x}{X}{\LinSigTyLimit{y}{Y(x)}{A\,x\,y}}
\)
has an inverse.
\end{axiom}

@@

\begin{itemize}
\item In \cref{fig:spf-act}, we provide the functorial actions of the $\el$
  and $\map$ operations used to define the indexed inductive types in
  \cref{fig:iilt}.
\item In \cref{fig:small-nonlin-ty,fig:small-lin-ty} we give the inference rules
  for the smallness judgements on nonlinear types and linear types, respectively.
\item In \cref{fig:full-non-linear-types}, we include the full set of rules for
  non-linear types, and in \cref{fig:jdg-eq-nonlinear} we provide their
  judgmental equalities.
\item In \cref{fig:jdgeq}, we give the judgmental equalities between
  linear terms in \theoryabbv.
\end{itemize}

\subsection{Non-linear Types}
\label{subsec:nonlinty}
We define sum types, list types, and $Fin~n$ from primitve non-linear types.

>>>>

% -*- fill-column: 80; -*-
\newif\ifarxiv
\arxivfalse

\PassOptionsToPackage{prologue,dvipsnames}{xcolor}

@@

%% Bibliography style
\bibliographystyle{ACM-Reference-Format}
%% Citation style
%% Note: author/year citations are required for papers published as an
%% issue of PACMPL.
\citestyle{acmauthoryear}   %% For author/year citations
%% \citestyle{acmnumeric}
\usepackage{mathpartir}
\usepackage{tikz-cd}
\usepackage{lipsum}
\usepackage{enumitem}
\usepackage{wrapfig}
\usepackage{fancyvrb}
\usepackage{xspace}
\usepackage{thmtools} % needed for sane cleveref naming of theorems
\usepackage{hyperref}
\usepackage[capitalise]{cleveref}

@@

\ifarxiv
\title{Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent
  Lambek Calculus (Extended Version)}
\else
\title{Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus}
\fi

@@

\ifarxiv\else {Proofs and syntactic forms elided from this paper are presented
    in full in the extended version of this paper \cite{schaefer2025intrinsicverificationparsersformal}.}\fi

@@

In addition to these type-theoretic principles, we need two additional
axioms that do not generally hold in systems based on linear
logic. First, we need that additive conjunction \emph{distributes}
over additive disjunction --- e.g., in the finitary case that $0 \& A
\cong 0$ and $(A + B) \& C \cong (A \& C) + (B \& C)$. More generally, we assume
\cref{ax:dist}.
% Comment from shepherd
% - The pagebreaks and figures make the explanation of axiom 3.1 a
% little confusing to read (with the text "In its most general form,"
% separated from the axiom it is intended to introduce by a pagebreak
% and a figure). You could fiddle with the layout a bit, or just revise
% that sentence to "More generally, we assume Axiom 3.1.", or similar.
%
% - If I understand it correctly, axiom 3.1 itself is using "e" as a variable name, which is confusing since it's also used as a metavariable for terms - I suggest picking a different variable name.
\begin{axiom}[Distributivity]
\label{ax:dist}
For any $A : (x : X) \to Y(x) \to L$, the definable distributivity function
%%%%%%%%%% An attempt to inline
% \(
%  {\LinSigTy{f}{\left(\PiTy{x}{X}{Y(x)}\right)}{\LinPiTy{x}{X}{A\,x\,(f\,x)}} \lto
% \LinPiTy{x}{X}{\LinSigTy{y}{Y(x)}{A\,x\,y}} \right)}
% \)
\(
 \lamblto{b}{\letin{\sigma\,f\,b'}{b}{\withlamb{x}{b'\,.\pi\,x}}} : \LinSigTyLimit{f}{\PiTy{x}{X}{Y(x)}}{\LinPiTyLimit{x}{X}{A\,x\,(f\,x)}} \lto
\LinPiTyLimit{x}{X}{\LinSigTyLimit{y}{Y(x)}{A\,x\,y}}
\)
has an inverse.
\end{axiom}

@@

\begin{itemize}
\item In \cref{fig:spf-act}, we provide the functorial actions of the $\el$
  and $\map$ operations used to define the indexed inductive types in
  \cref{fig:iilt}.
\item In \cref{fig:small-nonlin-ty,fig:small-lin-ty} we give the inference rules
  for the smallness judgments on nonlinear types and linear types, respectively.
\item In \cref{fig:full-non-linear-types}, we include the full set of rules for
  non-linear types, and in \cref{fig:jdg-eq-nonlinear} we provide their
  judgmental equalities.
\item In \cref{fig:jdgeq}, we give the judgmental equalities between
  linear terms in \theoryabbv.
\end{itemize}

\subsection{Non-linear Types}
\label{subsec:nonlinty}
We define sum types, list types, and $Fin~n$ from primitive non-linear types.
